---
title: "Vector Deepdive"
author: "Chantal Fuchs"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: flatly
engine: knitr
---

# Solutions for Vector Deepdive

::: {.callout-note}
**Declaration of AI Use:** Parts of this submission were developed with the
assistance of Claude (Anthropic). The AI was used to help with code structure, debugging, and R syntax. All results were verified by me.
:::

---

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(sf)
library(dplyr)
library(tidyr)
library(tictoc)
library(ggplot2)
library(knitr)
library(duckdb)

# Data paths
path_tlm   <- "Data/SWISSTLM3D_2026_LV95_LN02.gpkg"
path_bound <- "Data/swissBOUNDARIES3D_1_5_LV95_LN02.gpkg"
path_db    <- "Data/wald-kantone.duckdb"
path_chess <- "Data/chessboard.gpkg"

# Layer names
layer_cantons <- "tlm_kantonsgebiet"
layer_forest  <- "tlm_bb_bodenbedeckung"
```

---

## Task 1 – Forest percentage per canton (`sf`)

The goal is to find out how much of each Swiss canton is covered by forest.
I use the `swissTLM3D` dataset for the forest polygons and `swissBOUNDARIES3D`
for the canton boundaries, then clip the forest to each canton and calculate
the percentage.

### Step 1: Load canton boundaries

```{r}
#| label: sf-step1

tictoc::tic(msg = "sf | Step 1: Load canton boundaries")

cantons <- sf::st_read(path_bound, layer = layer_cantons, quiet = TRUE)
cantons <- sf::st_zm(cantons)  # drop Z dimension

tictoc::toc(log = TRUE)

cat("CRS:     ", st_crs(cantons)$epsg, "\n")
cat("Cantons: ", nrow(cantons), "\n")
```

### Step 2: Load and filter forest polygons

```{r}
#| label: sf-step2

tictoc::tic(msg = "sf | Step 2: Load and filter forest polygons")

forest <- sf::st_read(path_tlm, layer = layer_forest, quiet = TRUE)
forest <- sf::st_zm(forest)
forest <- forest[forest$objektart == "Wald", ]

tictoc::toc(log = TRUE)

cat("Forest polygons:", nrow(forest), "\n")
```

### Step 3: Align CRS

```{r}
#| label: sf-step3

tictoc::tic(msg = "sf | Step 3: Align CRS")

cat("Canton CRS:", st_crs(cantons)$epsg, "\n")
cat("Forest CRS:", st_crs(forest)$epsg,  "\n")

if (st_crs(cantons) != st_crs(forest)) {
  forest <- sf::st_transform(forest, st_crs(cantons))
}

tictoc::toc(log = TRUE)
```

### Step 4: Spatial intersection

```{r}
#| label: sf-step4

tictoc::tic(msg = "sf | Step 4: Intersect forest with cantons")

suppressWarnings({
  forest_canton <- sf::st_intersection(
    forest,
    cantons["name"]
  )
})

forest_canton$forest_area_m2 <- as.numeric(sf::st_area(forest_canton))

tictoc::toc(log = TRUE)

cat("Intersection rows:", nrow(forest_canton), "\n")
```

### Step 5: Calculate forest percentage per canton

```{r}
#| label: sf-step5

tictoc::tic(msg = "sf | Step 5: Calculate forest % per canton")

cantons$canton_area_m2 <- as.numeric(sf::st_area(cantons))

forest_sum <- forest_canton |>
  sf::st_drop_geometry() |>
  group_by(name) |>
  summarise(forest_area_m2 = sum(forest_area_m2, na.rm = TRUE), .groups = "drop")

result_sf <- cantons |>
  sf::st_drop_geometry() |>
  select(name, canton_area_m2) |>
  left_join(forest_sum, by = "name") |>
  mutate(
    forest_area_m2 = replace_na(forest_area_m2, 0),
    forest_pct     = (forest_area_m2 / canton_area_m2) * 100
  ) |>
  arrange(desc(forest_pct))

sf_total <- tictoc::toc(log = TRUE)

cat(sprintf("sf total time: %.2f s\n",
    sum(sapply(tictoc::tic.log(format=FALSE), function(x) x$toc - x$tic))))
```

### Results — Table

```{r}
#| label: sf-table

kable(
  result_sf |>
    mutate(
      `Canton Area (km²)` = round(canton_area_m2 / 1e6, 1),
      `Forest Area (km²)` = round(forest_area_m2 / 1e6, 1),
      `Forest (%)`        = round(forest_pct, 2)
    ) |>
    select(Canton = name,
           `Canton Area (km²)`,
           `Forest Area (km²)`,
           `Forest (%)`),
  caption = "Forest coverage per Swiss canton (sf)"
)
```

### Results — Bar chart

```{r}
#| label: sf-barplot
#| fig-width: 10
#| fig-height: 6

ggplot(result_sf,
       aes(x = reorder(name, forest_pct), y = forest_pct,
           fill = forest_pct)) +
  geom_col() +
  geom_hline(yintercept = mean(result_sf$forest_pct),
             colour = "tomato", linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = 2, y = mean(result_sf$forest_pct) + 1,
           label = sprintf("Swiss avg: %.1f%%", mean(result_sf$forest_pct)),
           colour = "tomato", size = 3.5) +
  scale_fill_gradient(low = "#d4edda", high = "#1a5c30", name = "Forest (%)") +
  coord_flip() +
  labs(
    title    = "Forest Coverage per Swiss Canton",
    subtitle = "Source: swissTLM3D 2026 × swissBOUNDARIES3D 1.5",
    x        = NULL,
    y        = "Forest cover (%)"
  ) +
  theme_minimal(base_size = 11)
```

### Results — Choropleth map

```{r}
#| label: sf-map
#| fig-width: 10
#| fig-height: 7

cantons_map <- cantons |>
  left_join(result_sf |> select(name, forest_pct), by = "name")

ggplot(cantons_map) +
  geom_sf(aes(fill = forest_pct), colour = "white", linewidth = 0.4) +
  scale_fill_gradient(low = "#d4edda", high = "#1a5c30", name = "Forest (%)") +
  geom_sf_text(aes(label = paste0(substr(name, 1, 3), "\n",
                                   round(forest_pct, 0), "%")),
               size = 2.2, colour = "black") +
  labs(
    title    = "Forest Cover per Canton",
    subtitle = "swissTLM3D 2026 × swissBOUNDARIES3D 1.5"
  ) +
  theme_void(base_size = 11) +
  theme(legend.position = "right")
```

---

## Task 2 – Forest percentage per canton (DuckDB)

Same analysis as Task 1, but using a pre-prepared DuckDB database
(`wald-kantone.duckdb`) instead of loading the raw files. The idea is to
see whether a SQL-based approach is faster.

```{r}
#| label: duckdb-setup

library(DBI)

con <- duckdb::dbConnect(duckdb::duckdb(), path_db, read_only = TRUE)
DBI::dbExecute(con, "INSTALL spatial; LOAD spatial;")

tables <- DBI::dbGetQuery(con, "SHOW TABLES")
cat("Tables in database:\n")
print(tables)

cat("\nwald columns:\n")
print(DBI::dbGetQuery(con, "DESCRIBE wald"))
cat("\nkantone columns:\n")
print(DBI::dbGetQuery(con, "DESCRIBE kantone"))
```

```{r}
#| label: duckdb-query

tictoc::tic(msg = "DuckDB | Full query: intersection + aggregation")

result_duckdb <- DBI::dbGetQuery(con, "
  WITH
    canton_areas AS (
      SELECT
        name                  AS canton_name,
        ST_Area(geom)         AS canton_area_m2,
        geom                  AS canton_geom
      FROM kantone
    ),
    forest_per_canton AS (
      SELECT
        c.canton_name,
        c.canton_area_m2,
        SUM(ST_Area(ST_Intersection(w.geom, c.canton_geom))) AS forest_area_m2
      FROM wald         AS w
      JOIN canton_areas AS c
        ON ST_Intersects(w.geom, c.canton_geom)
      GROUP BY c.canton_name, c.canton_area_m2
    )
  SELECT
    canton_name,
    ROUND(canton_area_m2  / 1e6, 2) AS canton_area_km2,
    ROUND(forest_area_m2  / 1e6, 2) AS forest_area_km2,
    ROUND(forest_area_m2 / canton_area_m2 * 100, 2) AS forest_pct
  FROM forest_per_canton
  ORDER BY forest_pct DESC
")

tictoc::toc(log = TRUE)
DBI::dbDisconnect(con)
```

### Results

```{r}
#| label: duckdb-table

kable(result_duckdb,
      col.names = c("Canton", "Canton Area (km²)",
                    "Forest Area (km²)", "Forest (%)"),
      caption = "Forest coverage per Swiss canton (DuckDB)")
```

### Execution time comparison

```{r}
#| label: timing-comparison

log_raw <- tictoc::tic.log(format = FALSE)

all_times <- sapply(log_raw, function(x) as.numeric(x$toc - x$tic))
all_names <- sapply(log_raw, function(x) x$msg)

sf_time     <- sum(all_times[!grepl("DuckDB", all_names)])
duckdb_time <- sum(all_times[grepl("DuckDB", all_names)])

cat(sprintf("sf total time    : %.2f s\n", sf_time))
cat(sprintf("DuckDB total time: %.2f s\n", duckdb_time))
if (duckdb_time > 0) {
  cat(sprintf("Speedup          : %.1fx\n", sf_time / duckdb_time))
}

data.frame(
  Approach = c("sf", "DuckDB"),
  time_sec = c(sf_time, duckdb_time)
) |>
  ggplot(aes(x = reorder(Approach, time_sec), y = time_sec, fill = Approach)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  scale_fill_manual(values = c("sf" = "#4a90d9", "DuckDB" = "#f5a623")) +
  coord_flip() +
  labs(title = "Execution Time: sf vs DuckDB",
       x = NULL, y = "Total time (seconds)") +
  theme_minimal()
```

### Discussion

I had no prior experience with either `sf` or DuckDB before this course, so
getting the code to run at all was already a challenge. One thing that surprised me early on was just how long loading the data takes — filtering the forest polygons alone took over 80 seconds, which I did not expect for what felt like
a simple step.

Looking at the results, Ticino stands out with over 42% forest cover. This makes sense geographically since it is a mountainous, heavily forested canton. On the other end, Geneva and Basel-Stadt are below 12%, which also makes sense as they are mostly urban. I had not expected the difference between cantons to be quite
so large.

For the timing comparison, DuckDB was about 2.4x faster than `sf`, 
which actually surprised me — I had expected the difference to be 
smaller since both tools use the same underlying GEOS library for 
the geometry calculations. Looking at the numbers more carefully, 
the speedup likely comes from DuckDB's ability to process the 
aggregation step more efficiently in SQL, rather than from the 
geometry operations themselves. With an even larger dataset, the 
advantage would probably be bigger since DuckDB can stream data 
without loading everything into memory at once, while `sf` has to 
hold it all in RAM.

---

## Task 3 – DE-9IM: King's case & Bishop's case

The DE-9IM (Dimensionally Extended 9-Intersection Model) is a way to describe the spatial relationship between two geometries. It looks at how the interior, boundary, and exterior of two shapes intersect, and encodes this as a 9-character string. For this task, I use it to distinguish between cells that share a full
edge (like a King's move) versus cells that only share a corner (like a Bishop's
move).

| | **I(B)** | **B(B)** | **E(B)** |
|:---|:---:|:---:|:---:|
| **I(A)** | `[0]` | `[1]` | `[2]` |
| **B(A)** | `[3]` | **`[4]`** | `[5]` |
| **E(A)** | `[6]` | `[7]` | `[8]` |

Symbols: `F` = empty, `0` = point, `1` = line, `2` = area, `*` = any.

The key position is `[4]` — the boundary-boundary intersection — which tells
us whether two cells share an edge (line = `1`) or only a corner (point = `0`).

### Load chessboard and compute DE-9IM strings

```{r}
#| label: chess-load
#| message: false
#| warning: false

centre_sf <- sf::st_read(path_chess, layer = "grid_orig", quiet = TRUE) |>
  sf::st_zm() |>
  sf::st_make_valid()

dest_sf <- sf::st_read(path_chess, layer = "grid_dest", quiet = TRUE) |>
  sf::st_zm() |>
  sf::st_make_valid()

sf::st_crs(centre_sf) <- NA
sf::st_crs(dest_sf)   <- NA

cat("Centre cells:", nrow(centre_sf), "\n")
cat("Dest cells:  ", nrow(dest_sf), "\n")

ggplot() +
  geom_sf(data = dest_sf,   fill = "lightgrey", colour = "black",
          linewidth = 1, inherit.aes = FALSE) +
  geom_sf(data = centre_sf, fill = "#2a9d8f",   colour = "black",
          linewidth = 1, inherit.aes = FALSE) +
  labs(title = "3×3 Chessboard",
       subtitle = "Green = centre cell (grid_orig)") +
  theme_void()
```

```{r}
#| label: chess-deim

rel_matrix <- sf::st_relate(centre_sf, dest_sf, sparse = FALSE)

results <- lapply(seq_len(nrow(dest_sf)), function(i) {
  deim_str <- rel_matrix[1, i]
  bb_dim   <- substr(deim_str, 5, 5)
  rel <- dplyr::case_when(
    bb_dim == "1" ~ "edge-adjacent",
    bb_dim == "0" ~ "corner-adjacent",
    TRUE          ~ "other"
  )
  data.frame(cell = i, `DE-9IM` = deim_str,
             `BB dim` = bb_dim, type = rel,
             check.names = FALSE)
})

df_deim <- do.call(rbind, results)
kable(df_deim, caption = "DE-9IM strings for each surrounding cell")

cat("\nEdge-adjacent cells (King):    ",
    df_deim$cell[df_deim$`BB dim` == "1"], "\n")
cat("Corner-adjacent cells (Bishop):",
    df_deim$cell[df_deim$`BB dim` == "0"], "\n")
```

### Manual derivation

#### (a) King's case – all adjacent cells (edge + diagonal)

**Edge-adjacent** cells share a full side → B(A)∩B(B) is a **line** (`1`):

```
        I(B)  B(B)  E(B)
I(A)  [  F  |  F  |  2  ]
B(A)  [  F  | *1* |  1  ]
E(A)  [  2  |  1  |  2  ]
```

→ DE-9IM = **`FF2F11212`**

- `I∩I = F`: the interiors of two touching polygons don't overlap.
- `B(A)∩B(B) = 1`: the shared edge is part of both boundaries → line.
- `I(A)∩E(B) = 2`: A's interior is entirely outside B → area.

**Corner-adjacent** cells only share a single point → B(A)∩B(B) is a **point** (`0`):

```
        I(B)  B(B)  E(B)
I(A)  [  F  |  F  |  2  ]
B(A)  [  F  | *0* |  1  ]
E(A)  [  2  |  1  |  2  ]
```

→ DE-9IM = **`FF2F01212`**

**King's case pattern** — covers both edge and corner neighbours using a wildcard:

→ **`FF2F*1212`**

#### (b) Bishop's case – diagonal cells only

The Bishop only moves diagonally, so only corner-adjacent cells count.
Position `[4]` must be `0` (point) to exclude edge-adjacent cells:

→ **`FF2F01212`**

### Visualisation

```{r}
#| label: chess-plot
#| fig-width: 10
#| fig-height: 5

edge_idx   <- df_deim$cell[df_deim$`BB dim` == "1"]
corner_idx <- df_deim$cell[df_deim$`BB dim` == "0"]

dest_sf$type <- dplyr::case_when(
  seq_len(nrow(dest_sf)) %in% edge_idx   ~ "Edge-adjacent (King)",
  seq_len(nrow(dest_sf)) %in% corner_idx ~ "Corner-adjacent (Bishop)",
  TRUE                                    ~ "Other"
)

p_king <- ggplot() +
  geom_sf(data = dest_sf,
          aes(fill = ifelse(type %in% c("Edge-adjacent (King)",
                                        "Corner-adjacent (Bishop)"),
                            type, "Other")),
          colour = "black", linewidth = 1) +
  geom_sf(data = centre_sf, fill = "#2a9d8f", colour = "black", linewidth = 1.5) +
  scale_fill_manual(
    values = c("Edge-adjacent (King)"     = "#e76f51",
               "Corner-adjacent (Bishop)" = "#f4a261",
               "Other"                    = "whitesmoke"),
    name = NULL
  ) +
  labs(title = "King's case", subtitle = "Edge + corner neighbours") +
  theme_void() +
  theme(legend.position = "bottom")

p_bishop <- ggplot() +
  geom_sf(data = dest_sf,
          aes(fill = ifelse(type == "Corner-adjacent (Bishop)",
                            type, "Other")),
          colour = "black", linewidth = 1) +
  geom_sf(data = centre_sf, fill = "#2a9d8f", colour = "black", linewidth = 1.5) +
  scale_fill_manual(
    values = c("Corner-adjacent (Bishop)" = "#f4a261",
               "Other"                    = "whitesmoke"),
    name = NULL
  ) +
  labs(title = "Bishop's case", subtitle = "Corner neighbours only") +
  theme_void() +
  theme(legend.position = "bottom")

library(patchwork)
p_king + p_bishop
```

### Summary table

| Case | Relationship | B(A)∩B(B) | DE-9IM |
|------|---|:---:|:---:|
| King (edge-adjacent) | share full edge | Line `1` | `FF2F11212` |
| King (corner-adjacent) | share corner point | Point `0` | `FF2F01212` |
| **King (combined)** | edge OR corner | `*` | **`FF2F*1212`** |
| **Bishop** | corner only | Point `0` | **`FF2F01212`** |

In PostGIS / DuckDB spatial:

```sql
-- King's case (all 8 neighbours)
WHERE ST_Relate(a.geom, b.geom, 'FF2F*1212')

-- Bishop's case (diagonal only)
WHERE ST_Relate(a.geom, b.geom, 'FF2F01212')
```


### Reflection

Task 3 was conceptually the most challenging part for me. The DE-9IM string
is not something I had encountered before, and it took a while to understand
what the 9 positions actually represent.

What helped was focusing on position [4] — the boundary-boundary intersection.
Once I understood that this single character is what distinguishes a shared
edge (line = `1`) from a shared corner (point = `0`), the rest followed
quite naturally.

What I found interesting is that the King and Bishop patterns share almost
the same DE-9IM string — the only difference is that one character. This
shows how precise the model is: it can distinguish two very similar spatial
relationships just from that one position.

If I were to use this in practice, I could see it being useful for things
like network analysis — for example, checking whether two road segments
truly share an edge or only touch at a point, which would affect routing
differently.